---
title: "HW 2"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F, R.options=list(max.print=100))

# Edit the file starting below
```

## Enya Liu el27773

**This homework is due on February 4, 2022 by 11:59pm. You will submit a pdf file on Gradescope.**

*For all questions, include the R code that you used to find your answer (show R chunk and outputs). Answers without supporting code will not receive credit (unless no code was required). Outputs without comments will not receive credit either: Write full sentences to describe your findings.*

### Question 1: (2 pts)

##### Type the word `letters` into the R chunk below. What does this predefined object in R contain? What is this object's data type/class? How many elements does it contain? *Include base R commands used to answer all three questions.*

```{r}
# test predefined word 'letters'
letters
# find type of 'letters'
class(letters)
# find how many elements 'letters' contains
length(letters)
```

'letters' contains lowercase letters from a to z, and all of the letters are characters.\
There are 26 characters in total.

------------------------------------------------------------------------

### Question 2: (2 pts)

##### Now, say we want to encode your name (or any other message) using a *cipher*. We want to replace every letter of a given character vector with the letter of the alphabet that is k positions after it in the alphabet. For example, if the letter was `a` and `k = 3`, we would replace it with `d`. We will also want it to loop around, so if the letter was `y` we'd replace it with `b`. For example, with `k = 3`, the word `dog` would become `grj`. You'll need a few tools to accomplish this. So let's take it step by step.

#### First, note that `A %in% B` tests each element of A to see whether it appears in B.

```{r}
letters %in% c("t","e","s","t")
```

##### How many elements are false in the resulting logical vector? *Include base R commands used to answer this question (not manually).*

```{r}
# calculate the true elements in logical vector
true_elements = sum(letters %in% c("t","e","s","t"))
# use total elements to minus the true elements to get the false elements
false_elements = length(letters) - true_elements
# get the result
false_elements
```

There are 23 false elements in the resulting logical vector.

------------------------------------------------------------------------

### Question 3: (2 pts)

##### Another function that will be useful is `which()`, a handy function that takes a logical vector argument and returns the indices/positions that are TRUE. For example, run the following code:

```{r}
# note: T is shorthand for TRUE, F for FALSE
which(c(F,T,F,T,F,T))
```

#### It tells you that elements in position 2, 4, and 6 are true. Now, use the `which` function, along with `%in%` and `letters` (from the previous questions), to find which positions in the alphabet the letters in the name `layla` occupy (saved as an object called name below). Do you notice any issue?

```{r}
# use which function to find the position of layla in letters
which(letters %in% c("l","a","y","l","a"))
```

The letters 'a', 'l', 'y' occupy in 1, 12, 25 positions of the alphabet.\
However, here is a issue that the position of repeated letters are not tested. So there is only 3 positions, and we do not know which position correspond to which letter. 

------------------------------------------------------------------------

### Question 4: (2 pts)

##### How can we avoid this? We can test each letter one at a time in their correct order! One approach would be to use a `for` loop. Write a for loop that goes through each element of the character vector `name` (i.e., `layla`) one at a time, finds its position in the alphabet, and saves each position in a vector called `positions`. Confirm that the positions are correct by using `positions` to find the corresponding letters with `letters`.

For example, the name "ali" would give you the positions 1,12, and 9. You can grab the letters in those positions by doing `letters[c(1, 12, 9)]`.

```{r}
# create an empty list to store the position
positions <- list()
# create a name like layla to test the for loop
name <- strsplit('layla', split = '')
# for loop iterates to find the position of letters
for (i in name) {
  positions <- match(i, letters)
  print(positions)
}
# check the position using the output of the for loop
letters[c(12,1,25,12,1)]
```

------------------------------------------------------------------------

### Question 5: (2 pts)

##### Instead of initializing the positions and writing a `for` loop for each new name we would like to encode, let's write a function! Define a function that takes a name (i.e., a character vector whose elements contain single letters) as the argument and returns the alphabetical positions each of those letters occupy. Call the function `get_position`. Once you have defined it, test it out with `layla`. Did you get all positions?

```{r}
# write a function
get_position <- function(name){
  for (i in name) {
  positions <- match(i, letters)
  }
  positions
}

# test the position
print(positions)

# test out the function with 'layla'
name <- strsplit('layla', split = '')
pos <- get_position(name)
letters[pos]
```

Yes, I got all positions for 'layla'

------------------------------------------------------------------------

### Question 6: (2 pts)

##### Let's encode the name `layla`! Shift all the positions by 1 and index `letters` to obtain the encoded name. Is that a real name?

```{r}
# find the position of 'layla'
# split the word
name <- strsplit('layla', split = '')
pos <- get_position(name)
# shift the positions by 1
pos = pos + 1
# encode with corresponding index
letters[pos]

```

The encoded name is 'mbzmb'. It is not a real name. 

------------------------------------------------------------------------

### Question 7: (3 pts)

##### What happens when we shift the positions past `z`, the 26th and final letter of the alphabet? Shifting the positions in `layla` up by k = 2 should give `ncanc`, but since there is no 27th element of letters, it will return `NA` instead of `a`. Try it in the code chunk below.

```{r}
# try a word with 'z'
name <- strsplit('zip', split = '')
pos <- get_position(name)
# shift the position
pos = pos + 1
# check the result
letters[pos]
# The result is NA  "j" "q" because there's no 27th letter of the alphabet. 

# try 'layla' up by k = 2
name <- strsplit('layla', split = '')
pos <- get_position(name)
# shift the position
pos = pos + 2
# check the result
letters[pos]
# The result is  "n" "c" NA  "n" "c" because there's no 27th letter of the alphabet. 

```

##### How do we make it loop around so that `z` shifted up 1 becomes `a`? In other words, how can we make 27 become 1, 28 become 2, 29 becomes 3, etc.? We will use the modulo operator `%%` (which tells you the remainder when you divide one number by another). Try running the code below, 27 %% 26 (pronounced "27 modulo 26") below. It returns 1, the remainder when the lefthand number (27) is divided by the right (26).

```{r}
27 %% 26
```

##### We just need our shifted positions modulo 26. You can do this with `(positions + k) %% 26`. One last minor issue: 26 %% 26 is 0 but we want it to return 26. There are a few fixes: 
- test if `positions + k` is less than 27: if it is, use `positions + k` for the encoded positions, otherwise use `positions + k %% 26`. 
- replace any 0 in the shifted vector by 26. 

##### Find some method you like to fix this issue. Then, put it all together: use the above techniques to encode the word `layla` by shifting every letter `k = 3` positions forward correctly. Is that a real name?

```{r}
# get 'layla' letters position
name <- strsplit('layla', split = '')
pos <- get_position(name)

# create an empty vector to store new positions
shift <- c()

# write a for loop with if statements to shift the position by 3
for (i in pos){
  if ((i + 3) < 27){
    i = i + 3
    shift <- c(shift, i)
  } else{
    shift <- c(shift, (i+3) %% 26)
  }
}

# check the result
letters[shift]
```

The result is 'odbod', and it is not a real name. 

------------------------------------------------------------------------

### Question 8: (2 pts)

##### One final objective: Write a function that incorporates all the work you have done to achieve this task. Name the function `cipher`. This function should take two arguments: a `word` (character vector to be encoded) and how many positions to shift (`k`). Fill in the code below with what you have been using above. Check your code with `layla` shifted by 3 positions and test your code with your own name with the shift of your choice! Is that a real name?

```{r}
# edit the code below (make sure to edit comment)

cipher <- function(word, k) {
  shifted <- c()
  position = get_position(list(word))
  for (i in position){
    if ((i + k) < 27){
      shifted <- c(shifted, i+k)
    } else{
      shifted <- c(shifted, (i+k) %% 26)
    }
  }
  letters[shifted]
}

# check
cipher(c("l", "a", "y", "l", "a"), 3)

# test your name!
cipher(c("e","n","y","a"), 3)
```

I checked the cipher function with 'layla' and it worked.\
Then I tested the function with my name 'Enya', and the output is 'hqbd' which is not a real name. 

------------------------------------------------------------------------

### Question 9: (1 pt)

##### A less guided question... You were given the code `oldp`. Can you decipher the code and find the name hidden behind it?

```{r}
# write a decipher function
decipher <- function(word, k) {
  shifted <- c()
  position = get_position(list(word))
  for (i in position){
    if ((i + k) > 0){
      shifted <- c(shifted, i+k)
    } else{
      shifted <- c(shifted, (i+k+26))
    }
  }
  print(letters[shifted])
}

# to find out the name, we need to find the index
# k is between -1 and -25
# test out all possibilities with a for loop
for (i in 1:25){
  decipher(c("o","l","d","p"),(-i))
}

```

Through out all possible combinations I found, the name hidden is 'liam'. 

------------------------------------------------------------------------

### Formatting: (2 pts)

##### Comment your code, write full sentences, and knit your file!


------------------------------------------------------------------------

    ```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```
