---
title: "HW 9"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F, R.options=list(max.print=100,dplyr.print_max=100))

# Edit the file starting below
```

## Enya Liu el27773

**This homework is due on April 22, 2022 by 11:59pm. You will submit a pdf file on Gradescope.**

*For all questions, include the R code that you used to find your answer (show R chunk and outputs). Answers without supporting code will not receive credit (unless no code was required). Outputs without comments will not receive credit either: Write full sentences to describe your findings.*

------------------------------------------------------------------------

###### Load the packages `tidyverse`, `factoextra`, and `cluster`.

```{r, message = FALSE}
# Call libraries (if factoextra or cluster were not installed, remember to install them)
library(tidyverse)
library(factoextra)
library(cluster)
```

------------------------------------------------------------------------

## Question 1: (2 pts)

###### The dataset for this homework comes from the article:

###### *Tsuzuku N, Kohno N. 2020. The oldest record of the Steller sea lion Eumetopias jubatus (Schreber, 1776) from the early Pleistocene of the North Pacific. <https://doi.org/10.7717/peerj.9709>*

###### Read the ***Abstract*** of the article and the section called *Results of Morphometric Analyses*. What was the goal of this study and what was the main finding?

The goal of the study is to identity the genus of the earliest sea lion fossil, GKZ-N 00001. \
The main finding was GKZ-N 00001 belongs to the Steller sea lion E. jubatus. 

------------------------------------------------------------------------

## Question 2: (1 pt)

###### Under the supplemental information, the data was retrieved from a word document into a `.csv` document. Import the dataset from GitHub. How many rows and how many columns are in this dataset? What does a row represent? What does a column represent?

```{r, message=FALSE}
library(tidyverse)

# upload data from github
HW9 <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//Sealions.csv")
```

```{r}
# take a look at this dataset
head(HW9)

```

There are 51 rowsand 39 columns.\
A row represent the data for one of the 51 sea lion sample.\
A column represent the sea lion sample, and other 38 columns represent measurements of external morphologies. 

------------------------------------------------------------------------

## Question 3: (1 pt)

###### Before we can analyze the data, let's do some cleaning. When importing this dataset into RStudio, which variables were considered numeric? Why are some measurements not considered as numeric?

```{r}
# use glimpse to check the class
glimpse(HW9)

```

K and AD were numeric variables. Some measurements were actually numeric variables but in a chr format because there are "-" in the columns. 

------------------------------------------------------------------------

## Question 4: (1 pt)

###### Using `dplyr` functions, replace all `-` in the dataset by missing values *NA* then make sure all measurements are defined as numeric variables with `mutate_at()`. Overwrite the dataset. What is the mean rostral tip of mandible `C`?

```{r}
# use mutate at to replace "-" with NA
HW9 <- HW9 %>%
  # vars select variables quickly and -XXX means deselect those variable
  mutate_at(vars(-ID, -K), ~na_if(.,"-")) 
# convert the variables to numeric variables
HW9[, c(2:39)] <- sapply(HW9[, c(2:39)], as.numeric)
# find mean of c
mean(HW9$C, na.rm = T) # ignore na
```
The mean rostral tip of mandible 'C' is 34.87


------------------------------------------------------------------------

## Question 5: (1 pt)

###### Using `dplyr` functions, only keep numeric variables that are not missing for the fossil specimen GKZ-N 00001. *Hint: you can use `select_if()` on the condition that `HW9_clean[51,]` has *no* missing value with `is.na()`).* Then remove the rest of the missing values and save the resulting dataframe as `HW9_NA`. How many columns and how many rows are remaining in this dataset?

```{r}
# your code goes below this line (make sure to edit comment)
HW9_NA <- HW9 %>%
  select_if(!(is.na(HW9[51,]))) %>%
  na.omit()
  

```

There are 42 rows and 23 columns remaining. 

------------------------------------------------------------------------

## Question 6: (1 pt)

###### Use `dplyr` functions on `HW9_NA` to split the `ID` variable into two variables `species` and `sex` with the function `separate()`. *Hint: in the `ID` variable, what symbol separates the species from sex?* The article states that the fossil specimen has to be male. Replace the missing value of `sex` for the fossil specimen GKZ-N 00001. *Hint: You could use the functions `mutate()` and `replace_na()`.* Save the resulting dataframe as `HW9_clean`.

```{r}
# your code goes below this line (make sure to edit comment)
HW9_clean <- HW9_NA %>%
  # here '\\' is needed because [ has a meaning in regex
  separate(col = ID, into = c("species", "sex"), sep = '\\[')
# only keep m or f as sex
HW9_clean$sex <- substr(HW9_clean$sex,0,1)
HW9_clean <- HW9_clean %>%
  # assign GKZ-N sex as male
  # replace na only in one column
  mutate(sex = replace_na(sex,"m" ))


```

------------------------------------------------------------------------

## Question 7: (1 pt)

###### Using `dplyr` functions, only keep numeric variables and scale each numeric variable. Save the resulting dataframe as `HW9_num`. What should the mean of the scaled variable of the rostral tip of mandible C be?

```{r}
# scale it
HW9_num <- HW9_clean %>%
  mutate_if(is.numeric,scale)
# find mean
mean(HW9_num$C)

```

The mean is 1.487009e-16 of C. 

------------------------------------------------------------------------

## Question 8: (2 pts)

###### Let's now perform PCA on the measurements available for the fossil specimen GKZ-N 00001. Using the function `prcomp()`, calculate the principal components (PCs) for the scaled data, `HW9_num`, obtained in Question 7. Construct a scree plot using the package `factoextra` with the function `fviz_eig()`. What is the cumulative percentage of explained variance for PC1 and PC2? 

```{r}
# visualize pca
HW9_num %>%
  select(is.numeric) %>%
  prcomp() %>%
  fviz_eig(addlabels = TRUE)

```

The cumulative percentage is 84.8% for just dimension 1. 

------------------------------------------------------------------------

## Question 9: (2 pts)

###### Using `fviz_contrib()`, find the 5 variables that contribute the most to PC1 and PC2. Comment on how differently these two PCs seem to be built.

```{r}
# visualize dm1 first
HW9_num %>%
  select(is.numeric) %>%
  prcomp() %>%
  fviz_contrib(choice = "var", axes = 1, top = 5)
# visualize dm2
HW9_num %>%
  select(is.numeric) %>%
  prcomp() %>%
  fviz_contrib(choice = "var", axes = 2, top = 5)
```

The top 5 variables that contribute most to dimension 1 are more evenly distributed.\
The top 1 variable to dimension 2 contributed 75%. 

------------------------------------------------------------------------

## Question 10: (2 pts)

###### How many *known species* are there in `HW9_clean` from question 6? Therefore, how many clusters should we look for to identify what species GKZ-N 00001 most likely belongs to? Using `pam()` on `HW9_num`, run the PAM clustering algorithm. *Note: make sure to call the `cluster` package.*

```{r}
library(cluster)
# apply pam
pam_results <- HW9_num %>%
  # here use k=3 because we have 3 species
  pam(k=3)
# Save cluster assignment as a column in your dataset
HW9_pam <- HW9_num %>%
  mutate(cluster = as.factor(pam_results$clustering))
HW9_pam
```

GKZ-N 00001 most likely belongs to species in cluster 1. 

------------------------------------------------------------------------

## Question 11: (2 pts)

###### Using `fviz_cluster()` and the PAM results in question 10, represent the observations. In the aesthetics, shape the observations by their `species`. The fossil specimen GKZ-N 00001 appears to be close to which species? *Note: make sure to add the legend for shape by adding `geom_point()` and `guides()`.*

```{r}
# your code goes below this line (make sure to edit comment)
fviz_cluster(pam_results, data = HW9_clean) +
  geom_point(aes(shape = HW9_clean$species)) +
  guides(shape = guide_legend(title = "shape"))

```

The GKZ-N 00001 appears to be close to E.jubatus species. 

------------------------------------------------------------------------

## Question 12: (2 pts)

###### Putting it all together. Reflect on and summarize in 1-2 sentences the different steps taken through this assignment. Compare your conclusions to the findings discussed by the researchers in the article (cite their findings).

First, we cleaned up the data by removing some columns and rows, then we scale it and did PCA and clustering to find which is the closest species to GKZ-N 00001.\
The researchers also find that GKZ-N 00001 is closest to E. jubatus species. 

------------------------------------------------------------------------

## Formatting: (2 pts)

###### Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```
