---
title: "HW 3 Key"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F, R.options=list(max.print=100))

# Edit the file starting below
```

## Enya Liu el27773

**This homework is due on February 11, 2022 by 11:59pm. You will submit a pdf file on Gradescope.**

*For all questions, include the R code that you used to find your answer (show R chunk and outputs). Answers without supporting code will not receive credit (unless no code was required). Outputs without comments will not receive credit either: Write full sentences to describe your findings.*

------------------------------------------------------------------------

### Question 1: (2 pts)

##### The dataset `ChickWeight` contains information about the weights (in grams) of chicks on four different diets over time (measured at 2-day intervals) as the result of an experiment. The first few observations are listed below.

```{r}
# Save dataset as a dataframe
ChickWeight <- as.data.frame(ChickWeight)

# Visuaize the first ten rows of the dataset
head(ChickWeight,10)
```

##### Use some combination of `table()` and `length()` to answer the following questions: 

```{r}
# find how many distinct chicks are there
length(table(ChickWeight$Chick))

# find how many distinct time points
length(table(ChickWeight$Time))

# find how many distinct diet conditions
length(table(ChickWeight$Diet))

# find how many chicks per diet condition
table(ChickWeight$Diet)

```
- How many distinct chicks are there?  
- How many distinct time points?  
- How many distinct diet conditions?  
- How many chicks per diet condition?  

- There are 50 distinct chicks.
- There are 12 time points.
- There are 4 distinct diet conditions.
- There are 220 chicks in 1 diet condition, 120 chicks in 2 diet condition, \
120 chicks in 3 diet condition, and 118 chicks in 4 diet condition. 

------------------------------------------------------------------------

### Question 2: (2 pts)

##### Using the `ggplot2` package, create a simple scatterplot showing chick `weight` (on the y-axis) as a function of `Time`. Label the axes including the units of the variables and give the plot a title. How does chick `weight` change over `Time`?

```{r}
# Call package
library(ggplot2)

# scatterplot of Time vs weight
ggplot(data = ChickWeight, aes(x = Time, y = weight)) + geom_point() + ggtitle('The Scatterplot of Time vs Weight') + xlab('Time (days)') + ylab('Weight(gm)')

```

As time increased, chick weight increased and more variations between individuals. 

------------------------------------------------------------------------

### Question 3: (2 pts)

##### Building upon the previous plot, map `Chick` to an aesthetic that assigns a color to each chick’s data points. Add lines that connect each chick’s points together with `geom_line()`. Finally, remove the legend. Do all chicks seem to gain weight in the same manner? Why/Why not?

```{r}
# create a scatterplot with colors that present different chicks
# remove the legends
ggplot(data = ChickWeight, aes(x = Time, y = weight, color = Chick)) + geom_point() + geom_line() + labs(title = 'Scatterplot of Weight by Time', x = 'Time(days)', y = 'Weight(gm)') + guides(color = FALSE, size = FALSE)

```

Not all chicks seem to gain weight in the same manner. Some chicks gain weight faster with a steeper slope on the scatterplot, while others have a smoother trend. This may be due to factors like diet. 

------------------------------------------------------------------------

### Question 4: (2 pts)

##### Continue modifying the same by removing the color from the points only (leave the lines colored by chick, but make all of the points white). Put the points *on top of* the lines. On which day was the last value of the chicks' weight recorded?

```{r}
# scatterplot with white points and colored lines
temp1 <- ggplot(data = ChickWeight, aes(x = Time, y = weight)) + geom_line(aes(color = Chick)) + geom_point(color = 'white') + labs(title = 'Scatterplot of Weight by Time', x = 'Time(days)', y = 'Weight(gm)')
# geom_point(aes(color = 'white')) turns out yellow points
# geom_point(color = 'white') turns out white points
# use aes(color = ?) when color = some variable, use color = ? when color is simply color
temp1
```
```{r}
# find which day was the last value of the chicks' weight
table(ChickWeight$Time)
```

On 21st day, the last value of chicks' weight was recorded. 

------------------------------------------------------------------------

### Question 5: (2 pts)

##### Now, facet this plot by diet. Can you tell from this new plot which diet results in greater weight? Explain.

```{r}
# 
temp1 + facet_wrap(~Diet) + labs(title = 'Scatterplot of Weight by Time', x = 'Time(days)', y = 'Weight(gm)')

```

All four plots have similar shapes. However, I would say diet 3 result in greater weight with a highest weight of about 360 gm at 21st day. Other diets only have highest weight below 350 gm. 

------------------------------------------------------------------------

### Question 6: (2 pts)

##### Go back to your plot from question 2 and fit a *linear regression line* (using `lm`) to the chicks in each diet with `geom_smooth()`. There should be 4 separate regression lines, one for each diet, each a separate color. Can you see more clearly which diet results in greater weight? Explain.

```{r}
# add linear regression line to each diet. 
ggplot(data = ChickWeight, aes(x = Time, y = weight)) + geom_point() + ggtitle('The Scatterplot of Time vs Weight') + xlab('Time (days)') + ylab('Weight(gm)') + geom_smooth(se = FALSE, method="lm", aes(color = Diet))

```

Diet 3 results in greater weight with a steeper slope. 

------------------------------------------------------------------------

### Question 7: (2 pts)

##### A scatterplot might not be the best way to visualize this data: it calls attention to the relationship between weight and time, but it can be hard to see the differences between diets. A more traditional approach for exploring the effect of diet would be to construct a barplot representing group means with standard error bars showing +/- 1 standard error.

##### Create a plot using `geom_bar` where each bar's height corresponds to the average chick weight for each of the four diet conditions. Rename the y-axis to include units (e.g., with scale_y_continuous(`name=`...)) and make the major tick marks go from 0 to 150 by 25 (e.g., with scale_y_continuous(`breaks=`...)). Which diet has the highest mean `weight`? 

```{r}
# barplot
ggplot(ChickWeight, aes(x = Diet, y = weight)) + stat_summary(geom='bar', fun.y='mean') + scale_y_continuous(name = 'Avg Weight (gm)', breaks = seq(0, 150, by =25))

```

Diet 3 has the highest mean weight. 

------------------------------------------------------------------------

### Question 8: (2 pts)

##### Add error bars showing + or - 1 SE using `geom_errorbar(stat = "summary")`. Make the error-bars skinnier by adding a `width =`**0.5** argument. Color the bars (not the error bars, but the barplot bars) by diet and change from the default color scheme using a `scale_fill_` or a `scale_color_`. diet seems to have the most variation in `weight`? The least variation?

```{r}
# add error bar with width = 0.5
# add color
ggplot(ChickWeight, aes(x = Diet, y = weight, fill = Diet)) + geom_bar(stat = 'summary', fun.y= mean) + scale_y_continuous(name = 'Avg Weight (gm)', breaks = seq(0, 150, by =25)) + geom_errorbar(stat = "summary", width = 0.5) + scale_fill_manual(values = c('blue','green','yellow','red'))


```

It seems like diet 3 has the most variation in weight, and diet 1 has the least variation in weight. 

------------------------------------------------------------------------

### Question 9: (2 pts)

##### Take your code from question 8 and replace `geom_bar()` with `geom_point()`. Remove the `breaks=` argument from `scale_y_continuous`. Make the points larger and color them all red. Put them *on top of* the error bars. Does the mean chick weight seem to differ based on the diet? *I am not asking to conduct hypothesis testing but informally state if they seem to differ and if so, how.*

```{r}
# change point size and color
ggplot(ChickWeight, aes(x = Diet, y = weight, fill = Diet)) + geom_errorbar(stat = "summary", width = 0.5) + geom_point(stat = 'summary', fun.y= mean, color = 'red', size = 3) + scale_y_continuous(name = 'Avg Weight (gm)') + scale_fill_manual(values = c('blue','green','yellow','red'))
```

Yes, the mean chick weight seem to differ based on diet. Diet 1 has the lowest mean chick weight and diet 3 has the highest mean chick weight. \
The reason for this is probably different amounts of food are consumed by chicks in different diet groups, so they grow differently in 21 days. 

------------------------------------------------------------------------

### Formatting: (2 pts)

##### Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```