---
title: "Worksheet 4: More visualization with ggplot"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
# The following code is a basic setup for your document
# You won't have to edit it (unless you want to!)
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F, R.options = list(max.print = 100))
```

## 1. Loading packages

Remember to call the package `ggplot2` which contains the functions we will use for visualization:

```{r}
# Call a package
library(ggplot2)
```

## 2. Getting a dataset

The dataset we will use for this worksheet is `msleep` which is part of the `ggplot2` package. It contains information about mammals, their sleeping habits, and other characteristics. Let's take a look:

```{r}
# Take a look at the dataset
head(msleep)

# More information about the observations, variables, ...
str(msleep)

# Let's save msleep as a data frame in our environment
msleep <- as.data.frame(msleep)
```

## 3. Representing categorical variables

### a. Using a bar graph

Let's investigate if mammals are mainly carnivore, herbivore, insectivore, or omnivore with a bar graph. What we would like to represent on the y-axis is the frequency of each category: that's a statistic!

```{r}
# Use geom_bar() with a statistic option
ggplot(msleep, aes(x = vore)) +
  geom_bar(stat = "count")
```

Note that using a `geom_` layer and a `stat` statement **or** using a `stat_` layer and a `geom` statement are equivalent:

```{r}
ggplot(msleep, aes(x = vore)) +
  stat_count(geom = "bar")
```

```{r}
# Replace `stat =` by setting aesthetics for the y axis, calculating a percentage
ggplot(msleep, aes(x = vore)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) 
```

### b. Handling missing values for a categorical variable

As showed in the first few plots, there is a category called NA, meaning that there is no information about the diet of the mammal.

```{r}
# Create a new dataset which excludes rows where vore is NA
msleep_nm <- msleep[!is.na(msleep$vore), ] 
# only select rows with non-missing values: ! and function is.na()
```

Now, let's visualize the distribution of the different diets with proportions, using labels, and adding some color by diet:

```{r}
# Now represent the diets without the NA category
ggplot(msleep_nm, aes(x = vore)) + 
  # fill will fill the bar, while color = '' only colors the outside of bars
  geom_bar(aes(y = (..count..)/sum(..count..), fill = vore)) + # fill the shape with some color
  labs(
    title = "Distribution of diets for mammals",
    x = "Types or diets",
    y = "Proportion"
  )
```

### c. Creating a new categorical variable

To differentiate between small and large mammals based on their body weight, let's create a new variable `size`.

```{r}
# Create a new variable with ifelse(condition, if_true, if_false)
# Use the median to separate the lower half of mammals (small) based on their body weight
msleep_nm$size <- ifelse(msleep_nm$bodywt < median(msleep_nm$bodywt), "Small", "Large")
```

```{r}
# Check the proportions: why does it make sense?
ggplot(msleep_nm, aes(x = size)) +
  geom_bar(aes(y = (..count..)/sum(..count..)))
```

### d. Comparing two categorical variables

We can create stacked plots that compares the distribution of two variables:

```{r}
# Let's represent the diets on the x-axis and the size as a color
ggplot(msleep_nm, aes(x = vore, fill = size)) +
  geom_bar()
```

Or design a side-by-side bar graph with the option `dodge`:

```{r}
# Side-by-side bar graph
ggplot(msleep_nm, aes(x = vore, fill = size)) +
  geom_bar(position = "dodge") 
```

Or with the relative distribution at each level (using proportions with `fill`):

```{r}
# Representing the distribution of size (color) for each diet (bar)
ggplot(msleep_nm, aes(x = vore, fill = size)) +
  geom_bar(position = "fill")
```

Note that the default label for the y-axis is count. We might want to overwrite that! Let's also switch the two variables:

```{r}
# Representing the distribution of diet (color) for each size (bar)
ggplot(msleep_nm, aes(x = size, fill = vore)) + 
  geom_bar(position = 'fill') +
  labs(y = "Proportion")
```

Instead of proportions, let's represent percentages by changing the scale of the y-axis:

```{r}
# Representing the distribution of diet (color) for each size (bar)
ggplot(msleep_nm, aes(x = size, fill = vore)) + 
  geom_bar(position = 'fill') +
  labs(y = "Percentage") + 
  # use an option called scales::percent to display percentages on tick marks for the y-axis
  scale_y_continuous(labels = scales::percent)
```

### e. Comparing one categorical variable with a numeric variable

We can represent the mean value of a numeric variable for each level of a categorical variable:

```{r}
# Represent the mean total amount of sleep depending on the diet
ggplot(msleep_nm, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_bar(stat = "summary", fun = "mean")
```

To represent how spread out the values are within a category, we can use error bars (representing +- 1 standard deviation from the mean):

```{r error}
ggplot(msleep_nm, aes(x = vore, y = sleep_total, fill = vore)) +
  geom_bar(stat = "summary", fun = "mean") +
  # adding error bar, here mean_se = mean standard error.
  # summary = create summary statistics
  geom_errorbar(stat = "summary", fun.data = "mean_se") 
```

### f. Dealing with many categories

If there are many categories, the bar plot can become messy because of the labels:

```{r}
# Many categories for order
ggplot(msleep_nm, aes(x = order, fill = order)) +
  geom_bar()
```

```{r}
# Solution 1 : Angle the axis labels and remove the legend
ggplot(msleep_nm, aes(x = order, fill = order)) +
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position="none") 
```

```{r}
# Solution 2 : flip the orientation of the bars
ggplot(msleep_nm, aes(y = order, fill = order)) +
  geom_bar() + 
  theme(legend.position="none")
```

## 4. Representing numeric variables

### a. Using a scatterplot with different shapes

Let's investigate the relationship between body weight and brain weight for different diets (represented by different shapes):

```{r}
# Use the aesthetics of shape
ggplot(msleep_nm, aes(x = bodywt, y = brainwt, shape = vore)) +
  geom_point(size = 3)
```

There are two data points that distort the view of the rest of the data points because the weights are much higher than other values:

```{r}
msleep_nm[msleep_nm$bodywt > 1000,]
```

### b. Changing scales

We can adjust the scale on the x-axis and y-axis to focus on the area where most of the points are:

```{r}
# Define new limits for the x- and y-axis
ggplot(msleep_nm, aes(x = bodywt, y = brainwt, shape = vore)) +
  geom_point(size = 3) +
  xlim(0,1000) + 
  ylim(0,2)
```

Because there is a high variation between the weights, we can consider a logarithmic scale for both the body weight and brain weight:

```{r}
# Adding a layer with `scale_`
ggplot(msleep_nm, aes(x = bodywt, y = brainwt, shape = vore)) +
  geom_point(size = 3) +
  # log helps us to visualize both small values and large values
  # better plot
  scale_x_log10() + 
  scale_y_log10()
```

```{r}
# Changing threshold for scientific notation
options(scipen = 999)

# Redo the graph
ggplot(msleep_nm, aes(x = bodywt, y = brainwt, shape = vore)) +
  geom_point(size = 3) +
  scale_x_log10() + 
  scale_y_log10()
```

### c. Playing with size, color, and shape

Let's compare the REM sleep associated with the total amount of sleep of these mammals for different diets, also taking into account their body weight.

```{r}
ggplot(msleep_nm, aes(x = sleep_total, y = sleep_rem, shape = vore, color = vore)) +
  geom_point(aes(size = bodywt))
```

Think about what the color, shape, size, represent. Is any information redundant?

Notice in the scatterplot that the variable `bodywt` does not seem to have many high values? Let's look at the distribution of body weight with a boxplot:

```{r}
# Distribution of `bodywt`
ggplot(msleep_nm, aes(y = bodywt)) +
geom_boxplot()
```

```{r}
# There are a few high values for body weight, let's try a log-transformation
ggplot(msleep_nm, aes(y = log10(bodywt))) +
geom_boxplot()
```

Now, let's compare REM sleep to the total amount of sleep for different levels of body weights in a logarithmic scale:

```{r}
# Use logarithmic scale of body weight for the size of the data points
ggplot(msleep_nm, aes(x = sleep_total, y = sleep_rem, color = log10(bodywt))) +
  geom_point(size = 3) +
  # choose colors of the scale from yellow to red
  scale_color_gradient(low = "yellow", high = "red")
```

## 5. Color palettes

The default colors created by ggplot can be replaced by calling some color palette. Here is a list of the different ones that exist:

```{r}
RColorBrewer::display.brewer.all()
```

A few examples with `scale_fill_brewer()`:

```{r}
# Shades of blue
ggplot(msleep_nm, aes(x = vore)) + 
  geom_bar(aes(y = (..count..)/sum(..count..), fill = vore)) +
  scale_fill_brewer()

# Shades of orange
ggplot(msleep_nm, aes(x = vore)) + 
  geom_bar(aes(y = (..count..)/sum(..count..), fill = vore)) +
  scale_fill_brewer(palette = "Oranges")
```
